The Impact of Disk Winds on the Optical Spectra of Cataclysmic Variables

### 2014-09-17 draft.tex
Lots of changes preparing for second draft

\begin{document}
%% define bibstyle and other definitions
\bibliographystyle{aabib}
%% renew commands
\renewcommand{\labelitemi}{$\bullet$}
\def\py{\textsc{Python} }
\def\tar{\textsc{Tardis} }
\def\cld{\textsc{Cloudy} }
\def\civ{C~\textsc{iv} }
\def\araa{ARAA}
\def\nat{Nature}
\def\apjl{ApJ Letters}
\def\aapr{AAPR}
\def\actaa{ACTAA} 
\def\ssr{SSR}
\def\apj{ApJ}
\def\pasp{PASP}
\def\aap{A\&A}
\def\mnras{MNRAS}
\def\aj{AJ}
\def\rmxaa{RMXAA}

@@

%\title{Modeling the Optical Spectra of Cataclysmic Variables: The Importance of Disk Winds}
\title{Disk Winds Matter: Simulating the Optical Spectra of Cataclysmic Variables}
\author{J. H. Matthews, C. Knigge, K. S. Long, S. A. Sim \& N. Higginbottom}
%%\altaffiltext{1}{School of Physics \& Astronomy, University of Southampton,
  %%Southampton, SO17 1BJ, UK}

@@

\begin{abstract}
High-state non-magnetic cataclysmic variables (CVs) 
exhibit strong hydrogen \& helium recombination lines in their optical 
spectra. These lines can be single peaked, and may be formed 
in the same bipolar outflow that is responsible for the
blue-shifted absorption lines seen in the ultraviolet (UV).
Here, we present results obtained by incorporating a `macro-atom' treatment into
our Monte Carlo radiative transfer code, which was
originally used to synthesize the UV spectra of disk-dominated CVs 
with a simple biconical wind. 
Using an identical model, we show that the same wind leaves a noticeable imprint 
on the optical spectrum, producing \ha\ and \heiiopt\ as well as
enhanced emission in the Ly-$\alpha$ and \heiiuv\ UV lines. 
In addition, our improved treatment of recombination means that the bound-free continuum
emission is significant, but does not yet overcome the intrinsic absorption edge
present in the stellar atmosphere models used to treat the accretion disk.
We find that a boundary layer is not essential for the production
of He II recombination lines, but also explore the effect of boundary layers
with temperatures of $80-200$kK on the ionization state of the wind
and resulting spectrum.
A short exploration of a 3-variable kinematic parameter space leads to models with higher densities 
that are able to produce single-peaked line emission, and also generate sufficient recombination emission
to fill in the Balmer jump photoabsorption edge.
Finally, we present a synthetic spectrum of a next generation model in and out of eclipse,
which shows a striking similarity with the optical spectrum of the high inclination
Nova-like variable RW Tri. The densities near the disk plane in this model
are such that \civ\ has a large collisionally excited component, in 
contrast to observations.
% We find that 
% a boundary layer is not essential to produce significant recombination to 
% He II, and if one is present we favour temperatures of $<10^5$K to avoid
% overionizing the UV resonance line species such as CIV. We also conduct Zanstra method
% calculations and estimate that one would infer a boundary layer temperature of around 80,000K
% from our benchmark calculation, despite it not being present in the model itself. This suggests that previous calculations
%done using this method are strongly dependent on both the assumed geometry and disk
%spectrum.
%We also note that
%the presence of a boundary layer becomes more noticeable in lower accretion rate systems.
\end{abstract}

@@

Outflows are ubiquitous  in accreting systems across many orders of magnitude in mass,
from young stellar objects \citep{bunn1995} right up 
to quasi-stellar objects \citep[QSOs;][]{ganguly2008}. Outflows can take the form of jets, such as those
seen in X-ray binaries \citep{fender2006} or active galactic nuclei \citep[AGN;][]{marscher2006}, 
but can also manifest as `winds' emanating from the accretion disk (REFS).
Understanding the impact of accretion disk winds on spectral features has far-reaching implications. 
They provide a feedback mechanism to limit the growth of black holes in AGN \citep{silkrees1998},
and their ever-present nature implies a deep connection with the accretion process. This connection
is important to understand in order to probe the universality of accretion \citep{mchardy2006, SKUK2012} 
and to determine the true extent to which
accretion disks conform to the `$\alpha$-disk' model proposed by 
\cite{shakurasunyaev1973}. 
Despite this, in many types of systems the impact of winds on observational
features is essentially unknown, and many theoretical questions relating to their origin 
and driving mechanism remain unanswered.

Cataclysmic variables (CVs) are systems in which a white dwarf accretes matter from a donor
star via Roche-lobe overflow. In non-magnetic, disk-dominated systems this accretion
is mediated by an accretion disk which forms around the white dwarf. 
nova-like (NL) variables are a particular subclass of objects in which the accretion disk
is in a permanent state of relatively high accretion rate 
($\dot{M} \sim 10^{-8}$~M$_{\odot}$~yr$^{-1}$),
making them the perfect laboratories to test the $\alpha$-disk model.
For over three decades, it has been known that winds emanating from the accretion disk
are important in shaping the ultraviolet (UV) spectra of high-state CVs (Heap 1978), 
the most spectacular evidence being the P-Cygni like profiles of resonance lines such as 
C${\textsc {iv}}$ (see e.g. Cordova \& Mason 1982\nocite{cordova1982}).

While the effect of the wind on the UV is at least partially clear, it may also have more subtle 
effects across the spectral range. It has been proposed that an accretion disk wind
can cause single-peaked emission line profiles in the optical (Murray \& Chiang 1996), 
such as those seen in Nova-like variables and, especially, the SW Sex stars (Honeycutt, Schlegel,
\& Kaitchuck 1986; Dhillon \& Rutten 1995). It is particularly telling that single-peaked 
lines in high-inclination systems can be seen even in eclipse, suggesting 
that a spatially extensive wind may be the source of emission (see figure~\ref{novalikes}).
%%(Honeycutt, Schlegel, \& Kaitchuck 1986; Dhillon \& Rutten 1995). 
Further evidence of the influence of winds in the optical can occasionally 
be seen in P-Cygni like signatures, observed in lines such as \ha\ 
and He \textsc{i} $\lambda5876$ \citep{RN98}.
In addition, \cite{KLWB98} suggested
that recombination continuum emission from a disk wind or atmosphere
could be responsible for filling in the Balmer absorption edge. This provides
a potential explanation for
the absence of a pronounced `Balmer jump' in optical spectra of CVs, which has never been adequately produced 
by theoretical models or spectral synthesis.
This dense disk atmosphere or wind transition region
necessary to produce the required recombination emission
could also explain the Fe~\textsc{ii} `curtain'
seen in near-UV observations of the high inclination dwarf nova OY Car in quiescence 
\citep{horne1994}.

@@

The first attempts to model disk winds in CVs applied techniques used in stellar
wind modeling to an accretion disk geometry (Mauche \& Raymond
1987; Drew 1987; Vitello \& Shlosman 1988). These efforts showed that mass loss rates of
$\sim10\%$ of the accretion rate
were required to produce the observed line strengths.
Using spectral synthesis codes to model resonance scattering,
Shlosman \& Vitello (1993; hereafter SV93) and Knigge, Woods \& Drew (1995; hereafter KWD95)
were both successful in reproducing the observed spectra such as the P-Cygni like
profiles in C\textsc{iv}~$\lambda1549$\AA.
Long \& Knigge (2002; LK02) presented a code that
solved self-consistently for thermal and ionization balance in the wind
according to the methods of Mazzali \& Lucy (1993; ML93). They successfully reproduced
UV resonance lines profiles and presented a synthesized spectrum of Z Cam
that agreed well with the observed features. Noebauer et al. (2010) used the same code 
to carry out more detailed comparisons to RW Tri and UX UMa. In addition to these 
radiative transfer calculations, some work has also been done using hydrodynamic wind models, 
such as that by \cite{pkdh2002}, and more recently by \cite{kusterer2014} with application to
AM Canum Venaticorum stars.

Here, we present work using the same Monte Carlo radiative transfer code as LK02, 
\py, but incorporating additional 
`macro-atom' techniques suggested by Lucy (2002, 2003). 
This improves our treatment of the Hydrogen and Helium
optical lines, as well as recombination continuum emission. 
Our goal is to investigate the effect of existing wind models
on the optical spectra of CVs.
In section 2 we present the code
and provide a brief description of macro-atoms and the radiative transfer methods we employ. 
We describe the kinematics and geometry of our biconical wind model in section 3.
In section 4, we present a benchmark model, which is discussed in section 5 with particular attention 
devoted to a small-scale parameter search. In section 6, we present a `next generation'
model which better reproduces the spectral features observed in NLs, and, 
finally, in section 7, we summarise our findings.

@@

The code is capable of operating in a number of different
regimes, both in terms of the scale of the system, and the 
shape and characteristic energies of the illuminating spectrum. 
It was originally described by LK02, who developed a biconical CV 
wind model to synthesize UV spectra. 
Higginbottom et al.\nocite{higginbottom2013} (2013; hereafter H13) used the same code to
present a benchmark model for broad absorption line (BAL) QSOs.
As part of this effort, the ionization scheme was improved so that 
the code is now capable of treating arbritrary SEDs.
%the power-law X-ray spectrum exhibited by AGN.
% This involved using a spectral model the mean intensity in a cell
% computed from MC estimators for, instead of using a dilute blackbody
% approximation. Compton processes
% were also incorporated into the heating and cooling balance.
In addition, \cite[][SDL05]{simmacro2005} modeled
Brackett and Pfund line profiles in YSOs using \textsc{Python,} which included
implementing a `macro-atom' mode in order to correctly treat the more
complex transition matrix involved in recombination.

@@

Previously, \py used a two-level atom approximation \cite[see e.g.][]{mihalas}. This approximation is 
fast, and works well as a first approximation when treating so-called `resonance lines' (such as C\textsc{iv}) 
in which the excited electron is strongly coupled to the ground state.
However, in a number of situations, for example, when a level is primarily populated from above, then
this becomes a poor representation. 
To reproduce recombination lines and lines in which the lower level is not the ground state, 
it is therefore important to use an improved line transfer technique and level population solver.

By quantising matter into `macro-atoms', and radiant and kinetic 
energy into indivisible energy packets (r- and k- packets respectively), 
Lucy (2002, 2003\nocite{lucy2002, lucy2003}; hereafter L02, L03) showed that it is possible 
to asymptotically reproduce the emissivity of a gas in statistical equilibrium 
without simplifying the treatment of line transfer. 
His macro-atom scheme allows all possible transition paths from a given level and a full non-Local 
Thermodynamic Equilibrium (NLTE) solution for the level populations based on MC estimators. 
For a full description consult L02 and L03.

Macro-atoms have been used to model astrophysical plasmas in 
Wolf-Rayet star winds \citep{sim2004}, AGN disk winds \citep{simlong2008, tatum2012}
and supernovae \citep{kasen2006, kerzendorfsim}.
SDL05 used \py itself to apply this technique to
a Hydrogen-only model. We are now able to adopt a hybrid scheme. 
Any species in which have their level populations determined by
more complex interactions than resonance transitions
can be treated as full macro-atoms. The other species are treated as so-called `simple-ions',
where they still follow the indivisible packet constraint but use the two-level atom approximation.
This allows one to preserve the fast treatment of resonance lines while providing 
an improved representation of, for example, recombination, which is important when modeling features 
such as the Balmer lines and also recombination {\sl continuum} emission. 
Recombination is expected to be an important process in the optical
spectra of CVs, and they thus make a perfect application for Lucy's 
advances in MC methods.

@@

In our hybrid ionization and excitation scheme, 
simple-ions have their ionization fractions
calculated using a modified Saha equation, and internal level populations are
calculated from LTE. For macro-atoms, we solve the equations of statistical 
equilibrium using MC estimators for the radiation field. This enables us to 
determine the ionization and excitation state of the ions in question.

\subsubsection{Macro-atoms}
% In order to correctly model lines in which the lower level is not the ground state 
% (such as H$\alpha$ \& H$\beta$), and improve the treatment of recombination,
% we treat H \& He as macro-atoms. This is because 
% the levels responsible for the line emission are populated mostly from above.
% Therefore, 
we treat H \& He as macro-atoms. 
Macro-atom NLTE level populations and ion fractions are calculated by solving 
the statistical equilbrium equations between each pair of levels. The rate
between a lower level $l$ and upper level $u$ in an ion is given by

\begin{eqnarray}
{\cal R}_{ul} &= & \beta_{lu} A_{ul} n_l + \beta_{lu} B_{ul} n_l J_{lu}^b + C_{ul} n_u n_e\label{eq:nlte_rul}\\
{\cal R}_{lu} &= & \beta_{lu} B_{lu} n_l J_{lu}^b + C_{lu} n_l n_e,
\end{eqnarray}

where $A$ and $B$ are Einstein coefficients, $C$ is the collisional rate coefficient and
$\beta_{lu}$ is the probability that a given line photon will escape the Sobolev
region. $J^b$ is the MC estimator for mean intensity at the blue wing of the line as 
recorded during the photon propagation. In the context of 
bound-free, we consider rates between the continuum state (or, in the case of ions with more than one 
bound electron, the ground state of the upper ion), $\kappa$,
and lower level $l$,

\begin{eqnarray}
{\cal R}_{\kappa l} &= & \alpha_{\kappa l} n_{\kappa} n_e \\
{\cal R}_{l \kappa} &= & n_l \int_{\nu_0}^{\infty} \frac{ 4 \pi J_{\nu} \sigma_{\nu}}{h \nu} d\nu,
\end{eqnarray}

where $\sigma_{\nu}$ is the photoioniziation cross section, $\alpha_{\kappa l}$ is the recombination
coefficient to level $l$ and $J_{\nu}$ is the mean intensity.
This treatment means that radiative and collisional rates to and from all 
levels are considered when calculating both the ionization state, and the level populations, 
although we neglect photoionization to excited levels of the upper ion. 
The \cite{vanregemorter} approximation is used for collisional transitions,
and thus we assume collisions between radiatively forbidden collisions are
negligible, a reasonable approximation in the regime where levels are mainly populated
by recombination. 
%{2002A&A...384..725L}.  source for beta sobolevs
%$\beta_{lu}=\frac{1}{\tau_{lu}}[1-\exp(-\tau_{lu})]$

@@

C, O, N, Ne, Na, Mg, Al, Si, S, Ar, Ca \& Fe are all included as two-level atoms. 
This is a first order approximation, which is appropriate 
when treating resonance lines in which the lower level is at or close to the 
ground state. 
For these elements, we follow LK02 in using the ionization scheme presented by Mazzali \& Lucy (1993). 
Specifically, we use the formula
\begin{equation}
\frac{n_{j+1} n_e}{n_j} = W [\xi + W(1-\xi)]
\left(\frac{T_e}{T_R}\right)^{1/2}
\left(\frac{n_{j+1}n_e}{n_j}\right)^*_{T_R} \label{ionization}
\end{equation}
to solve for the relative ionization fractions. In this equation, the `starred' term on 
the right represents the Saha abundances, in accordance with the notation of Mihalas (1982). 
$W$ is an effective dilution factor, $\xi$ is the
fraction of recombinations going directly to the ground state, and
$T_R$ and $T_e$ are the radiation and electron temperature
respectively. LK02 provide a full description of our implementation.

The reason for using this scheme 
is twofold. First, unlike in H13, the SED can be approximated well by a dilute blackbody due to the 
absence of a power-law X-ray source, and therefore the above modification to the Saha equation
is appropriate; Second, this provides continuity from LK02 and allows us to do a like-for-like comparison.
This is especially useful in assessing the improvement in Hydrogen line emission, and any changes
that might occur in the ionization stages of the resonance line species such as CIV. 
Excitation is then calculated in LTE. 
%This is because the 
% level populations are only important in working out the resonance scattering 
% opacity through the wind when modeling resonance lines.

\subsection{Atomic Data}
Atomic data is obtained from the same sources described in LK02 and updated in H13. 
The additional level and line information required to treat He as a macro-atom was 
obtained from \textsc{Topbase} \citep{topbase2005}. We treat H as a 20-level atom as in SDL05, where 
each level is defined by the principal quantum number, $n$, which is appropriate due to
the degeneracy of each level in the H atom. He~\textsc{II} is treated in much the same way,
but with 10 levels. However, He~\textsc{I} has larger energy differentials between different l subshells
and triplet and singlet states. Thus, we still include levels up to $n=10$ but explicitly 
treat the $l$ and $s$ sub-orbitals as distinct levels instead of assuming they are perfectly `l-mixed',
This also allows the model to produce singlet and 
triplet lines, which are seen in optical CV spectra \citep[e.g.][]{dhillon1996}.

@@

Here we describe some of the features of our biconical wind model. Our aim here is not
to provide a `best fit' model to match any particular dataset.
Rather, we take as a starting point
the types of models which have already been successful in explaining UV spectra of CVs.
We can then examine the contribution of the wind to 
the optical features when the treatment of line transfer
is improved.

@@

\section{Discussion of Results}

@@

\subsection{Exploring Parameter Space}

>>>>

\begin{document}
%% define bibstyle and other definitions
\bibliographystyle{aabib}
%% renew commands
\renewcommand{\labelitemi}{$\bullet$}
\def\py{\textsc{Python} }
\def\tar{\textsc{Tardis} }
\def\cld{\textsc{Cloudy} }
\def\civ{C~\textsc{iv} }
\def\araa{ARAA}
\def\nat{Nature}
\def\apjl{ApJ Letters}
\def\aapr{AAPR}
\def\actaa{ACTAA} 
\def\ssr{SSR}
\def\apj{ApJ}
\def\apss{AP\&SS}
\def\pasp{PASP}
\def\aap{A\&A}
\def\mnras{MNRAS}
\def\aj{AJ}
\def\rmxaa{RMXAA}

@@

%\title{Modeling the Optical Spectra of Cataclysmic Variables: The Importance of Disk Winds}
%\title{Disk Winds Matter: Simulating the Optical Spectra of Cataclysmic Variables}
\title{The Impact of Disk Winds on the Optical Spectra of Cataclysmic Variables}
\author{J. H. Matthews, C. Knigge, K. S. Long, S. A. Sim \& N. Higginbottom}
%%\altaffiltext{1}{School of Physics \& Astronomy, University of Southampton,
  %%Southampton, SO17 1BJ, UK}

@@

\begin{abstract}
High-state non-magnetic cataclysmic variables (CVs) 
exhibit strong hydrogen and helium recombination lines in their optical 
spectra. These lines can be single peaked, and may be formed 
in the same bipolar outflow that is responsible for the
blue-shifted absorption lines seen in the ultraviolet (UV).
Here, we present results obtained by incorporating a `macro-atom' treatment into
our Monte Carlo radiative transfer code, which was
originally used to synthesize the UV spectra of disk-dominated, high-state CVs 
with a simple biconical wind. 
Using an identical model, we show that the same wind leaves a noticeable imprint 
on the optical spectrum, producing \ha\ and \heiiopt\ as well as
enhanced emission in the Ly-$\alpha$ and \heiiuv\ UV lines. 
% In addition, our improved treatment of recombination means that the bound-free continuum
% emission is significant, but does not yet overcome the intrinsic absorption edge
% present in the stellar atmosphere models used to treat the accretion disk.
We find that a boundary layer is not essential for the production
of He II recombination lines.
%  but also explore the effect of boundary layers
% with temperatures of $80-200$kK on the ionization state of the wind
% and resulting spectrum.
A short exploration of a 3-variable kinematic parameter space leads to models with higher densities 
that are able to produce single-peaked line emission, and also generate sufficient recombination emission
to fill in the Balmer jump photoabsorption edge intrinsic to our disk atmosphere spectra.
Finally, we present a synthetic spectrum of an improved optical model in 
and out of eclipse, which shows a striking similarity with 
the optical spectrum of the high inclination
Nova-like variable RW Tri. The densities near the disk plane in this model
are such that \civfull\ has a large collisionally excited component, in 
contrast to observations.
% We find that 
% a boundary layer is not essential to produce significant recombination to 
% He II, and if one is present we favour temperatures of $<10^5$K to avoid
% overionizing the UV resonance line species such as CIV. We also conduct Zanstra method
% calculations and estimate that one would infer a boundary layer temperature of around 80,000K
% from our benchmark calculation, despite it not being present in the model itself. This suggests that previous calculations
%done using this method are strongly dependent on both the assumed geometry and disk
%spectrum.
%We also note that
%the presence of a boundary layer becomes more noticeable in lower accretion rate systems.
\end{abstract}

@@

% Outflows are ubiquitous  in accreting systems across many orders of magnitude in mass,
% from young stellar objects \citep{bunn1995} right up 
% to quasi-stellar objects \citep[QSOs;][]{ganguly2008}. Outflows can take the form of jets, such as those
% seen in X-ray binaries \citep{fender2006} or active galactic nuclei \citep[AGN;][]{marscher2006}, 
% but can also manifest as `winds' emanating from the accretion disk (REFS).
% Understanding the impact of accretion disk winds on spectral features has far-reaching implications. 
% They provide a feedback mechanism to limit the growth of black holes in AGN \citep{silkrees1998},
% and their ever-present nature implies a deep connection with the accretion process. This connection
% is important to understand in order to probe the universality of accretion \citep{mchardy2006, SKUK2012} 
% and to determine the true extent to which
% accretion disks conform to the `$\alpha$-disk' model proposed by 
% \cite{shakurasunyaev1973}. 
% Despite this, in many types of systems the impact of winds on observational
% features is essentially unknown, and many theoretical questions relating to their origin 
% and driving mechanism remain unanswered.

Cataclysmic variables (CVs) are systems in which a white dwarf accretes matter from a donor
star via Roche-lobe overflow. In non-magnetic, disk-dominated systems this accretion
is mediated by an accretion disk which forms around the white dwarf. 
Nova-like variables (NLs) are a particular subclass of objects in which the accretion disk
is in a permanent state of relatively high accretion rate 
($\dot{M} \sim 10^{-8}$~M$_{\odot}$~yr$^{-1}$),
making them the perfect laboratories to test the widely-used 
`$\alpha$-disk' model proposed by 
\cite{shakurasunyaev1973}.

For over three decades, it has been known that winds emanating from the accretion disk
are important in shaping the ultraviolet (UV) spectra of high-state CVs (Heap 1978), 
the most spectacular evidence being the P-Cygni like profiles of resonance lines such as 
\civfull\ (see e.g. Cordova \& Mason 1982\nocite{cordova1982}).
Despite this, in many types of systems the impact of winds on observational
features is essentially unknown, and many theoretical questions relating to their origin 
and driving mechanism remain unanswered.

While the effect of the wind on the UV is at least partially clear, it may also have more subtle 
effects across the spectral range. It has been proposed that an accretion disk wind
can provide a mechanism for single-peaked emission lines (Murray \& Chiang 1996), 
such as those seen in Nova-like variables and, especially, the SW Sex stars \citep{HSK86, DR95}.
It is particularly telling that single-peaked 
lines in high-inclination systems can be seen even in eclipse, suggesting 
that a spatially extensive wind may be the source of emission (see figure~\ref{novalikes}).
%%(Honeycutt, Schlegel, \& Kaitchuck 1986; Dhillon \& Rutten 1995). 
Further evidence of the influence of winds in the optical can occasionally 
be occasionally seen in P-Cygni like profiles observed in lines such as \ha\ 
and He \textsc{i} $\lambda5876$ \citep{RN98}.

% Somewhere, probably the introduction, we need to discuss alternatives for
% explaining the Balmer Jump problem. The other major possibility is that
% the disk atmospheres are not correct, see e.g Shaviv or that
% backscattering fromt he wind changes the disk structure. This is the main
% weakness in our model approach, because we cannot calculate the atmosphere
% ourselves. (Christian and I have had some discussions with Hillier, who is
% here in Aspen, and is the author of CMFGEN.  There was a paper on which he
% was a co-author where 1-d models of winds were summed to ³make² a CV wind.
% These used CMFGen and to calculate the structure of the wind, and so may
% include the backr]reaction of the wind on the disk photosphere.  He said
% he would check if these models had Balmer jumps, but you might want to
% find the paper).

It has also been proposed
that recombination continuum emission from a disk wind or `transition region'
could be responsible for filling in the Balmer absorption edge  \citep{KLWB98}. 
This provides a potential explanation for
the absence of a pronounced `Balmer jump' in optical spectra of CVs, 
which has never been adequately produced by using stellar atmospheres 
to model the accretion disk (see e.g. Haug 1987)\nocite{haug1987}. 
However, it is also possible that accretion disks do not possess
and intrinsic deep Balmer absorption edges.
This depends to the optical depth at the edge
and can change due to decrease due to less temperature variation with height
near the disk photosphere (see e.g. Shaviv et al. 1991; 
but also more recent work by Idan et al. 2010).
\nocite{idanshaviv2010} \nocite{shaviv1991}

% This dense disk atmosphere or wind transition region
% necessary to produce the required recombination emission
% could also explain the Fe~\textsc{ii} `curtain'
% seen in near-UV observations of the high inclination dwarf nova OY Car in quiescence

% In addition, \cite{KLWB98} suggested
% that recombination continuum emission from a disk wind or atmosphere
% could be responsible for filling in the Balmer absorption edge. This provides
% a potential explanation for
% the absence of a pronounced `Balmer jump' in optical spectra of CVs, which has never been adequately produced 
% by theoretical models or spectral synthesis.
% This dense disk atmosphere or wind transition region
% necessary to produce the required recombination emission
% could also explain the Fe~\textsc{ii} `curtain'
% seen in near-UV observations of the high inclination dwarf nova OY Car in quiescence 
% \citep{horne1994}.

@@

The first attempts to model disk winds in CVs applied techniques used in stellar
wind modeling to an accretion disk geometry (Mauche \& Raymond
1987; Drew 1987; Vitello \& Shlosman 1988). These efforts showed that mass loss rates of
$\sim10\%$ of the accretion rate
were required to produce the observed line strengths.
Using spectral synthesis codes to model resonance scattering,
Shlosman \& Vitello (1993; hereafter SV93)\nocite{SV93} and \cite{KWD95}
were both successful in reproducing the observed spectra such as the P-Cygni like
profiles in C\textsc{iv}~$\lambda1549$\AA.
Long \& Knigge (2002; hereafter LK02)\nocite{LK02} presented a code that
solved self-consistently for thermal and ionization balance in the wind
according to the methods of Mazzali \& Lucy (1993; ML93)\nocite{ML93}. They successfully reproduced
UV resonance lines profiles and presented a synthesized spectrum of Z Cam
that agreed well with the observed features. Noebauer et al. (2010) used the same code 
to carry out more detailed comparisons to RW Tri and UX UMa. In addition to these 
radiative transfer calculations, some work has also been done using hydrodynamic wind models, 
such as that by \cite{pkdh2002}, and more recently by \cite{kusterer2014} with application to
AM Canum Venaticorum stars.

Here, we present work using the same Monte Carlo radiative transfer code as LK02, 
but incorporating additional `macro-atom' techniques suggested by Lucy (2002, 2003). 
Our aim is to model high-state CVs over a wide spectral range.
In particular, we hope to assess the potential contribution
to the optical spectrum from a disk wind model which preserves the UV features.
It will then be possible to see how closely
our models resemble the spectra in figure 1 as we change wind parameters.
We will investigate whether a wind can produce significant
recombination emission lines and whether the recombination
continuum emission is sufficient to mask the Balmer jump from a disk atmosphere.
We will also learn if the mechanism for single-peaked line emission
proposed by \cite{MC96} works with plausible model parameters.
In section 2 we present the code and provide a brief description of macro-atoms 
and the radiative transfer methods we employ. 
We describe the kinematics and geometry of our biconical wind model in section 3.
In section 4, we present a benchmark model, which is explored on in section 5 
with a small-scale parameter search. In section 6, we present a
model which better reproduces the optical spectrum of a NL, and, 
finally, in section 7, we summarise our findings.

% Here, we present work using the same Monte Carlo radiative transfer code as LK02, 
% \py, but incorporating additional 
% `macro-atom' techniques suggested by Lucy (2002, 2003). 
% This improves our treatment of the Hydrogen and Helium
% optical lines, as well as recombination continuum emission. 
% Our goal is to investigate the effect of existing wind models
% on the optical spectra of CVs.

@@

The code is capable of operating in a number of different
regimes, both in terms of the scale of the system, and the 
shape and characteristic energies of the illuminating spectrum. 
It was originally developed by LK02 in order to model the UV spectra
of CVs with a simple biconical disk wind model.. 
Higginbottom et al.\nocite{higginbottom2013} (2013; hereafter H13) used the same code to
present a benchmark model for broad absorption line (BAL) QSOs.
As part of this effort, the ionization scheme was improved so that 
the code is now capable of treating arbritrary SEDs.
%the power-law X-ray spectrum exhibited by AGN.
% This involved using a spectral model the mean intensity in a cell
% computed from MC estimators for, instead of using a dilute blackbody
% approximation. Compton processes
% were also incorporated into the heating and cooling balance.
In addition, Sim et al. (2005; hereafter SDL05)\nocite{simmacro2005} modeled
Brackett and Pfund line profiles in young-stellar objects (YSOs) using \textsc{Python,} which included
implementing a `macro-atom' mode in order to correctly treat recombination lines.

@@

Previously to SDL05, \py used a two-level atom approximation \cite[see e.g.][]{mihalas}. This approximation is 
fast, and works well as a first approximation when treating so-called `resonance lines' (such as \civfull) 
in which the excited electron is strongly coupled to the ground state.
However, in a number of situations, for example, when a level is primarily populated from above, then
this becomes a poor representation. 
To reproduce recombination lines and lines in which the lower level is not the ground state, 
it is therefore important to use an improved line transfer technique and level population solver.

By quantising matter into `macro-atoms', and radiant and kinetic 
energy into indivisible energy packets (r- and k- packets respectively), 
Lucy (2002, 2003\nocite{lucy2002, lucy2003}; hereafter L02, L03) showed that it is possible 
to `asymptotically reproduce the emissivity of a gas in statistical equilibrium 
without simplifying the treatment of line transfer'. 
His macro-atom scheme allows all possible transition paths from a given level and a full non-Local 
Thermodynamic Equilibrium (NLTE) solution for the level populations based on MC estimators. 
For a full description consult L02 and L03. 
Macro-atoms have been used to model astrophysical plasmas in 
Wolf-Rayet star winds \citep{sim2004}, AGN disk winds \citep{simlong2008, tatum2012},
supernovae \citep{kasen2006, kerzendorfsim} and YSOs (SDL05).
% Recombination is expected to be an important process in the optical
% spectra of CVs, and they thus make a perfect application for Lucy's 
% advances in MC methods.

@@

Whereas SDL05 used \py itself to apply the macro-atom 
technique to a Hydrogen-only model, we
adopt a hybrid ionization and excitation scheme. 
Species can be treated as either full macro-atoms
or using the two-level atom approximation.
In the latter case, they still follow the indivisible packet constraint.
This allows one to preserve the fast treatment of resonance lines while providing 
an improved representation of processes with more complex
transition paths such as recombination. This is important when modeling features 
such as the Balmer lines and also recombination {\sl continuum} emission.

% In our hybrid ionization and excitation scheme, 
% simple-ions have their ionization fractions
% calculated using a modified Saha equation, and internal level populations are
% calculated from LTE. For macro-atoms, we solve the equations of statistical 
% equilibrium using MC estimators for the radiation field. This enables us to 
% determine the ionization and excitation state of the ions in question.

\subsubsection{Macro-atoms}
% In order to correctly model lines in which the lower level is not the ground state 
% (such as H$\alpha$ \& H$\beta$), and improve the treatment of recombination,
% we treat H \& He as macro-atoms. This is because 
% the levels responsible for the line emission are populated mostly from above.
% Therefore, 
We treat H and He as macro-atoms. 
Macro-atom NLTE level populations and ion fractions are calculated by solving 
the statistical equilbrium equations between each pair of levels. The bound-bound 
de-excitation rate, ${\cal R}_{ul}$, in an ion is given by

\begin{equation}
{\cal R}_{ul} = \beta_{lu} A_{ul} n_u + \beta_{lu} B_{ul} n_u J_{lu} + C_{ul} n_u n_e,\label{eq:nlte_rul}
\end{equation}

where $u$ and $l$ denote the upper and lower levels, $C$ is the collisional rate coefficient
and $A$ and $B$ are the Einstein coefficients. $\beta_{lu}$ is the probability that a 
given line photon will escape the Sobolev region. $J$ is the MC estimator for the mean intensity in the line as 
recorded during the photon propagation. The corresponding excitation rate is then

\begin{equation}
{\cal R}_{lu} = \beta_{lu} B_{lu} n_l J_{lu} + C_{lu} n_l n_e.
\end{equation}

We also calculate the photoionization rates between a lower level $l$,
and the continuum state (or, in the case of ions with more than one 
bound electron, the ground state of the upper ion), $\kappa$,

\begin{equation}
{\cal R}_{l \kappa}= n_l \int_{\nu_0}^{\infty} \frac{ 4 \pi J_{\nu} \sigma_{\nu}}{h \nu} d\nu,
\end{equation}

where $\sigma_{\nu}$ is the photoionization cross section and $J_{\nu}$ is the mean intensity. 
The recombination rate is given by

\begin{equation}
{\cal R}_{\kappa l} = \alpha_{\kappa l} n_{\kappa} n_e, \\
\end{equation}

where $\alpha_{\kappa l}$ is the recombination
coefficient to level $l$. 
This treatment means that radiative and collisional rates to and from all 
levels are considered when calculating both the ionization state, and the level populations, 
although we neglect photoionization to excited levels of the upper ion. 
The \cite{vanregemorter} approximation is used for collisional transitions,
and thus we assume collisions between radiatively forbidden transitions are
not important in determining level populations.
This is a reasonable approximation in the regimes considered here.
%{2002A&A...384..725L}.  source for beta sobolevs
%$\beta_{lu}=\frac{1}{\tau_{lu}}[1-\exp(-\tau_{lu})]$

@@

Lines from all other elements are treated with the two-level atom approximation. 
This is a first order approximation, which is appropriate 
when treating resonance lines in which the lower level is at or close to the 
ground state. 
For these elements, we follow LK02 in using the ionization scheme presented by Mazzali \& Lucy (1993). 
Specifically, we use the formula
\begin{equation}
\frac{n_{j+1} n_e}{n_j} = W [\xi + W(1-\xi)]
\left(\frac{T_e}{T_R}\right)^{1/2}
\left(\frac{n_{j+1}n_e}{n_j}\right)^*_{T_R} \label{ionization}
\end{equation}
to solve for the relative ionization fractions. In this equation, the `starred' term on 
the right represents the Saha abundances, in accordance with the notation of Mihalas (1982). 
$W$ is an effective dilution factor, $\xi$ is the
fraction of recombinations going directly to the ground state, and
$T_R$ and $T_e$ are the radiation and electron temperature
respectively. LK02 provide a full description of our implementation.
This ionization scheme is appropriate here because
the SED can be approximated well by a dilute blackbody due to the 
absence of a power-law X-ray source, unlike with AGN (see e.g. H13).
For two-level atoms, we assume LTE level populations.
% This ionization scheme is appropriate here because, unlike with AGN,
% the SED can be approximated well by a dilute blackbody due to the 
% absence of a power-law X-ray source, and therefore the above modification to the Saha equation
% is appropriate; Second, this provides continuity from LK02 and allows us to do a like-for-like comparison.
% This is especially useful in assessing the improvement in Hydrogen line emission, and any changes
% that might occur in the ionization stages of the resonance line species such as CIV. 
% Excitation is then calculated in LTE. 
% This is because the 
% level populations are only important in working out the resonance scattering 
% opacity through the wind when modeling resonance lines.

\subsection{Atomic Data}
Here we use the same atomic data as LK02 and updated by H13.
We have obtained the additional level and line information required to treat He as a macro-atom 
from \textsc{Topbase} \citep{topbase2005}. We treat H as a 20-level atom as in SDL05, where 
each level is defined by the principal quantum number, $n$, which is appropriate due to
the degeneracy of each level in the H atom. He~\textsc{II} is treated in much the same way,
but with 10 levels. However, He~\textsc{I} has larger energy differentials between different l subshells
and triplet and singlet states. Thus, we still include levels up to $n=10$ but explicitly 
treat the $l$ and $s$ sub-orbitals as distinct levels instead of assuming they are perfectly `l-mixed',
This also allows us to model singlet and 
triplet lines, which are seen in optical CV spectra \citep[e.g.][]{dhillon1996}.

@@

Here we describe some of the features of our biconical wind model.

@@

\section{Exploring Parameter Space}

@@
